---
title: "p8105_hw2"
author: "yl6378"
date: "2026-10-07"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Problem 1
```{r}
subway_df = read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv"
) |>
  select(
    Line,
    `Station Name`,
    `Station Latitude`,
    `Station Longitude`,
    Route1:Route11,
    `Entrance Type`,
    Entry,
    Vending,
    ADA
  ) |>
  mutate(
    Entry = if_else(Entry == "YES", TRUE, FALSE)
  )
```

After cleaning, the subway dataset contains 1,868 observations and 19 variables. The retained variables include subway line and station information, station latitude and longitude, routes served, entrance type, entry status, vending availability, and ADA accessibility. I also converted the `Entry` variable from `"YES"` / `"NO"` to a logical variable. At this stage, the data are not fully tidy because route information is spread across multiple columns (`Route1` through `Route11`) rather than stored in a single route variable.

```{r}
subway_df |>
  distinct(Line, `Station Name`) |>
  nrow()
```

There are 465 distinct subway stations.

```{r}
subway_df |>
  filter(ADA == TRUE) |>
  distinct(Line, `Station Name`) |>
  nrow()
```

There are 84 ADA-compliant subway stations.

```{r}
subway_df |>
  filter(Vending == "NO") |>
  summarize(
    total = n(),
    entry_yes = sum(Entry),
    prop_entry = entry_yes / total
  )
```

Among station entrances and exits without vending machines, 69 out of 183 allow entrance. This corresponds to a proportion of approximately 37.7%.

```{r}
subway_long = subway_df |>
  mutate(
    across(Route1:Route11, as.character)
  ) |>
  pivot_longer(
    Route1:Route11,
    names_to = "route_number",
    values_to = "route_name",
    values_drop_na = TRUE
  )
```

The route variables are reshaped from wide to long format. The original `Route1` through `Route11` columns are converted into two variables: `route_number`, which indicates the original route column, and `route_name`, which contains the subway route served.

```{r}
subway_long |>
  filter(route_name == "A") |>
  distinct(Line, `Station Name`) |>
  nrow()
```

There are 60 distinct subway stations that serve the A train

```{r}
subway_long |>
  filter(route_name == "A", ADA == TRUE) |>
  distinct(Line, `Station Name`) |>
  nrow()
```

Among the stations that serve the A train, 17 are ADA compliant

# Problem 2

# Problem 3